



# Week 6: Reshape


## Package `reshape`

Package `reshape` provides a unified approach to aggregation, based on
an extended formula notation.  The core idea is to create a "melted"
version of a dataset using the `melt` function, which can then be cast
(using function `cast`) into an object with the desired orientation.
This means that we want to rearrange the form (not the content) of the
data to suit the need of analysis.

Before using the package, we need to familiarize the "anatomy" of a
dataset. We started by thinking of data interms of vector, matrix, and
data frame.  In a data frame, observations are in rows and variables
are in columns.  To reshape a dataset, we can classify variables into
"identifier" and "measured" variables. 

An identifier (`id`) identifies the unit that the measurements take
place.  Id variables are usually discrete. A measured variable
represents what is measured on that unit, typically the response or
predictor variables. 

For example, in the `iris` dataset, we have fout measured variables
(`Sepal.Length, Sepal.Width, Petal.Length, Petal.Width`) and one id
variables (`Species`).  Later we reshaped the data to have three
columns, `Value, Variable`, and `Species`.  In this reshaped dataset,
the column `Value` is the measured variable and `Variables, Species`
are id variables.  In other words, we can generalize the data
classification to one measured variable and the rest are id variables.

### Melting data
The dataset with one column of measured varaible values and id
variables is the most basic data form which we can generate new
forms. 

To produce the basic form of dataset is not a simple task.  The
process of creating it is called melting.  The `reshape` package
provides the function `melt` to carry out the task.  For example,  
```{r}
iris.melt <- melt(iris, id = "Species",
                  measured=c("Sepal.Width", "Sepal.Length", "Petal.Length","Petal.width"))
head(iris.melt)
```
The only assumption when using `melt` is that all measured variables
are in the same type (e.g., numeric, factor, date).

The dataset `iris.melt` is now a molten data, ready to be casted into
different forms.  In some cases, your datasets are already in the
molten form.  All you need to do is to make sure that the value column
is named `value`. 

### Casting molten data
We now have a molten dataset and we can cast it into different forms.
All we need is to tell the function `cast` what form we need using a
formula. 

The casting formula describes the shape of the output format.  The
general form of a formula includes two groups of variables separated
by a tilde.  Variables on the left-hand side of the tilde are column
variables and those on the right-hand side are row variables.  For
example,  

```{r}
cast(iris.melt, Species ~ variable, length)
cast(iris.melt, Species ~ variable, mean)
```
Sumamrizes the sample size and mean.

The function `dcast` casts a molten data into data frame format.  The
minimum input are (1) a molten data and (2) a formula.  The formula
can also take $\cdot$ (a single dot) and $\cdots$ (three dots) to
represent no variable and all other variables (not already included in
the formula).

```{r}
cast(iris.melt, Species ~ variable)
cast(iris.melt, ... ~ variable)
cast(iris.melt, Species+variable ~ .)
```

What if we want to return to the original data frame?  In this case,
the molten data lacks one piece of information:
```{r}
iris$rep <- rep(1:50, 3)
iris.melt2 <- melt(iris, id=c("Species", "rep"),
                   measured=c("Sepal.Width", "Sepal.Length", "Petal.Length","Petal.width"))
cast(iris.melt2, Species+rep ~ variable)
```

### The French Fries Example

```{r}
ffm <- melt(french_fries, id=1:4, na.rm=T)

### Counting nonmissing values:
cast(ffm, .~., length)

### counting by treatment
cast(ffm, treatment ~ . , length)
cast(ffm, . ~ treatment, length)

### rep by treatment
cast(ffm, rep ~ treatment)

### treatment by rep:
cast(ffm, treatment ~ rep)

cast(ffm, treatment + rep ~ .)
cast(ffm, rep + treatment ~ .)
cast(ffm, . ~ rep+treatment)
```

Summary:
```{r}
cast(ffm, rep+treatment~variable)
cast(ffm, rep+treatment~variable, mean)
cast(ffm, rep+treatment~variable, sd)
cast(ffm, treatment ~ ., c(mean, sd))
cast(ffm, treatment ~ ., summary)
cast(ffm, treatment ~ variable + result_variable, c(mean, sd))
cast(ffm, treatment + variable ~ result_variable, c(mean, sd))
```

Conditional operator

```{r}
cast(ffm, treatment~rep|variable, mean)
```
## Package `reshape2`

Reshape2 is a reboot of the reshape package, a new package for
reshaping data that is much more focused and     faster.

This version improves speed at the cost of functionality.

What's new in reshape2:

-- considerably faster and more memory efficient thanks to a much
better underlying algorithm that uses the power and speed of
subsetting to the fullest extent, in most cases only making a single
copy of the data. `cast`is replaced by two functions depending on the
output type: dcast produces data frames, and acast produces
matrices/arrays. 

-- multidimensional margins are now possible: grand_row and grand_col
have been dropped: now the name of the margin refers to the variable
that has its value set to (all).

-- some features have been removed such as the | cast operator, and
the ability to return multiple values from an aggregation function.

-- a new cast syntax which allows you to reshape based on functions of
variables (based on the same underlying syntax as plyr): better
development practices like namespaces and tests.


## Reshaping Data
THe advantage of using R for statistical analysis is that R comes with
functions to carry out statistical analysis.  But each function
requires a specific input format.  As a result, reshaping data to suit
for the need of a function is the most commonly performed task. In
most cases, R functions require input using data frame.  We will focus
on working with data frames.

### Modifying data frame variables

Data frames are lists.  New variables can be creates by simply
assigning their value to a column that doesn't already exist in the
data frame.  Let's use the loblolly pine growth data as an example.
This dataset has three columns recording the growth of loblolly pine
trees over time.
```{r}
head(Loblolly)
```
Variable transformation is probably the most common modification to a dataset.  We add the transformed variable into an additional column:
```{r}
Loblolly$logheight <- log(Loblolly$height)
```
(When we use data from R or R packages, all changes we made will be
kept in the memory of the current R session.  Data from the system
won't be changed.) 

When the data frame name is long, we often use the function
`transform` to reduce the amount of typing.   
```{r}
Loblolly <- transform(Loblolly, logheight=log(height))
```

We access variables in a data frame through the data frame name plus
the `$` operator and variable names.  The function `with` helps to
reduce the repeated typing of the data frame name.  When using `with`,
R puts the data frame temperarily on the top of the search list: 
```{r}
logheight<-with(Loblolly, log(height))
```
When this line is submitted, R looks into the data frame `Loblolly` first to find `height`.  If a wrong name was typed, R would not be able to find it in the data frame, R will continue to search outside the data frame.
```{r}
height2 <- runif(20, 3, 85)
logheight2 <- with(Loblolly, log(height2))
```
We can remove a column of a data frame by setting its values to be `NULL`:
```{r}
head(Loblolly)
Loblolly$logheight <- NULL
head(Loblolly)
```
When a similar operation is needed for several columns and we want to
overwrite the original versions, we can put the multiple columns we
want to replace on the left hand side of the assignment operator.  For
example, if we want to convert the unit of iris measurements from
centimeters to inches, we need to divide each of the 4 measurement
variables by 2.54: 
```{r}
iris[,-5] <- sapply(iris[,-5], function(x) x/2.54)
```

Recoding a variable is another form of data transformation.  For
example, when the actual value of sepal length is not important, but
whether the length is above or below 6 cm is important, we would like
to have a column to indicate whether sepal length is above 6 or
not. This is a simple logic comparison: 
```{r}
largesepal <- iris$Sepal.Length > 6
```
In some cases, the function `ifelse` is very handy. The function takes
a logical vector as its first argument and two other arguments: the
first provides a value for the case where elements of the input
logical vector are true, and the second for the case where they are
false.  For example, suppose that we have experiments carried out in
10 sites and we realized that sites 2 and 4 can be grouped in one
group and the rest in another group.  The grouping can be created as
follows: 
```
newgroup <- ifelse(sites %in% c(2,4), 1, 2)
```
To make thing more complicated, suppose that sites 2 and 4 are in
group 1, sites 5 and 8 are in group 2 and the rest are in group 3: 
```
newgroup <- ifelse(sites %in% c(2,4), 1, 
                   ifelse (sites %in% c(5, 8), 2, 3))
```

### Reshaping a data frame
Not all data frame have all the necessary identifiers.  For example,
the `iris` data frame lacks the flower indentifier.  Let's work with
the iris data for one species: 
```{r}
iris.setosa <- iris[iris$Species=="setosa",-5]
head(iris.setosa)
```
Many R functions for statistical analysis requires that the data frame be reshaped to have a column of measured value and a column of variable.  
```{r}
setosaM <- melt(iris.setosa,
                measured=c("Sepal.Width", "Sepal.Length", "Petal.Length","Petal.width")) 
```
In this case, a much simpler approach would be to use the function `stack`:
```{r}
setosaS <- stack(iris.setosa)
```

The function `unstack` returns the data to its original form:
```{r}
setosaUS <- unstack(setosaS, values~ind)
```

### Function `reshape`
Not to be confused with the package `reshape`, the function `reshape`
is written for data with repeated measures to convert a data frame
between long and wide forms.  A long form data frame puts multiple
measurements of a subject (replicates) in separate rows, while the
wide form puts them in the same row.  Here is an example of long form: 
```{r}
obs <- data.frame(subj = rep(1:4, each = 3), 
                  time=rep(1:3, 4),
                  x=rnorm(12), y=rnorm(12))
head(obs)
```
A wide format has 4 rows, one for each subject:
```{r}
obs.wide <- reshape(obs, idvar="subj", v.names=c("x","y"), 
                         timevar="time", direction="wide")
obs.wide
str(obs.wide)

reshape(obs.wide, idvar="subj", direction="long")
```

### Combining Data Frames

The two frequently used functions are `rbind` (binding data frames by
row) and `cbind` (binding by column). 
```{r}
x <- data.frame(a=c("A","B","C"), b=c(1,2,3))
y <- data.frame(a=c("D","E","F","G"), b=c(3,4,5,6))

z <- rbind(x, y)
```
When using `rbind`, the two data frames must have the same names.  When using `cbind`, the two data frames must have the same number of rows (or one is a multipler of the other):
```{r}
cbind(y, z=c(1,2))
cbind(y, z=rep(1:2, 2))
```
But not:
```
cbind(x, z=c(1,2))
```

Merging two data frames based on a common factor is often necessary.
The function `merge` provides a flexible means for merging. 
```{r}
x<-data.frame(a=c(1,2,4,5,6), x=c(9,12,14,21,8))
y<-data.frame(a=c(1,3,4,6), y=c(8, 14, 19, 2))
merge(x, y)
```
The result may not be what you expect.  

By default, `merge` will find the common name in `x` and `y` to use as
the variable for merging (argument `by`).  If the merging variable has
different names in `x` and `y`, we can use argument `by.x` and `by.y`.
The argument `all` (default `all=FALSE`) is a logical value indicating
whether all rows in both `x` and `y` should be included: 
```{r}
merge(x, y, all=T)
```
If only all rows of `x` (`y`) are needed, we specify `all.x=T` (`all.y=T`).
```
merge(x, y, all.x=T)
merge(x, y, all.y=T)
```

One more example of `merge`:
```{r}
authors <- data.frame(
    surname = I(c("Tukey", "Venables", "Tierney", "Ripley", "McNeil")),
    nationality = c("US", "Australia", "US", "UK", "Australia"),
    deceased = c("yes", rep("no", 4)))
books <- data.frame(
    name = I(c("Tukey", "Venables", "Tierney",
             "Ripley", "Ripley", "McNeil", "R Core")),
    title = c("Exploratory Data Analysis",
              "Modern Applied Statistics ...",
              "LISP-STAT",
              "Spatial Statistics", "Stochastic Simulation",
              "Interactive Data Analysis",
              "An Introduction to R"),
    other.author = c(NA, "Ripley", NA, NA, NA, NA,
                     "Venables & Smith"))

m1 <- merge(authors, books, by.x="surname", by.y="name")
m2 <- merge(books, authors, by.x = "name", by.y = "surname")
m3 <- merge(books, authors, by.x = "name", by.y = "surname", all=T)
```

### Merging using index
The function `merge` is convenient.  In many cases, especially when we want to mearge data by a set of unique index (e.g., site names), we can simply use subscripting.  For example:
```{r}
x <- data.frame(site=c("S3","S2","S1"), Ag=c(20, 31, 19))
y <- data.frame(site=c("S1","S1","S1","S2","S2","S3","S3","S1","S2","S3"),
                y=rnorm(10))
```
Because the site names are the same, we can first sort `x` by site:
```{r}
x <- x[order(x$site),]
x
```
then:
```{r}
oo <- as.numeric(ordered(y$site))
oo
y$Ag <- x$Ag[oo]
y
```


