

# Week 3

## Data Management Objectives
- Subsetting data using square brackets
- Using `tapply`
- Using logic comparisons



- Florida is always a problem state, not only in presidential
    election
- Setting TP criterion for the Everglades -- almost a decade long
    litigation 
- Setting nutrient criteria for coastal waters -- another long
    legal battle 
- Legal battles between EPA and the State, between the State and
    environmental groups, between State and industry, between
   environmental groups and EPA, ...

## Background -- EPA's Approach 

Criteria for monitoring station inclusion
- Stations must have > 4 observations of TN, TP, Chla, and Turbidity
- All 0 concentrations observations should be dropped
- At least 2 years of data
- Annual means must be based on $> 4$ data points, at least 1 in
    summer and 1 in winter
- Once the data are cleaned, calculate annual geometric means
    for each station

## Using Preprocessed Data


All existing data were pooled and stations near each other are combined
- The resulting data `SouthFlorida.csv`
- Data summary
```{r, echo=FALSE}
sf_path=paste(dataDIR, "SouthFlorida.csv", sep="/")
Data <- read.csv(sf_path, header=T)
summary(Data)[,c(5,21)]
```

## Data Processing Considerations

- Drop sites with fewer than 4 observations
- Drop sites with fewer than 4 data points for TP, TN, CHLA, or Turbidity 
- Drop observations with 0 concentration values for TP, TN, CHLA, or Turb
- Drop sites with less than 2 years of records
- Drop sites without summer or winter observations

## Data for Analysis

After processing, calculating annual geometric means of TP, TN, CHLA, and Turb

R Coding Considerations

- Dates -- summer and winter
- Counting by site
- Calculating log-means for by site-year combination
- EPA used function `tapply`

```{r}
#tapply(Data, INDEX, FUN, ..., simplify=TRUE)
```

## tapply


### Step 1: Dropping sites with fewer than 4 observations (rows)

- Find sites with fewer than 4 TP samples

```{r}
site.drop <- tapply(Data$TP.S, Data$Site, FUN=function(x) sum(!is.na(x))<4)
```

- The function used in `tapply` is `sum(!is.na(x)) < 4`: returns a logic value of whether the total number of non-missing is less than 4 
- The results is a vector of `TRUE` and `FALSE`, each is associated with a site name


```{r}
sites.drop <- tapply(Data$TP.S, Data$Site,
      function(x) sum(!is.na(x)) < 4) |
              tapply(Data$TN.S, Data$Site,
      function(x) sum(!is.na(x))<4)   |
              tapply(Data$TURB.S, Data$Site,
      function(x) sum(!is.na(x))<4)   |
              tapply(Data$CHLA, Data$Site,
      function(x) sum(!is.na(x))<4)
sites.drop <- names(sites.drop)[sites.drop]
if(length(sites.drop)>0)
    Data <- Data[!is.element(Data$Site, sites.drop),]
```

### Step 2: Dropping 0 concentration values
```{r}
Data$TP.S[Data$TP.S==0] <- NA
Data$TN.S[Data$TN.S==0] <- NA
Data$TURB.S[Data$TURB.S==0] <- NA
Data$CHLA[Data$CHLA==0] <- NA
```

### Step 3: Dropping Years with Fewer Than 4 Observations

```{r}
site.yr <- paste(Data$Site, Data$Year)
siteyr.drop <-
  tapply(Data$CHLA, site.yr, 
        function(x) sum(!is.na(x))<4) |
  tapply(Data$TURB.S, site.yr, 
        function(x) sum(!is.na(x))<4) |
  tapply(Data$TP.S, site.yr, 
        function(x) sum(!is.na(x))<4) |
  tapply(Data$TN.S, site.yr, 
        function(x) sum(!is.na(x))<4)
siteyr.drop <- 
  names(siteyr.drop)[siteyr.drop]
if(length(siteyr.drop)>0)
  Data <- Data[!is.element(site.yr, 
                     siteyr.drop),]
```

### Step 4: At Least One Obs in Summer and One in Winter

Define summer and winter using dates

- Convert date column into a `Date` object

```{r}
Data$dateR <- as.Date(Data$DateT)
```

- Extract month:

```{r}
Data$Month <- ordered(months(Data$dateR), 
                      levels=month.name)
Data$Quarter <- quarters(Data$dateR)
tmp <- as.numeric(Data$Month)
Data$Summer <- 0
Data$Summer[tmp>=5 & tmp<=9] <- 1
## EPA forgot to code winter
```

Use `tapply` in Step 4

```{r}
siteyr.summ <- paste(Data$Site, Data$Year, Data$Summer)
siteyrsum.drop <-
  tapply(Data$CHLA, siteyr.summ, 
         function(x) sum(!is.na(x))==0) |
  tapply(Data$TURB.S, siteyr.summ, 
         function(x) sum(!is.na(x))==0) |
  tapply(Data$TP.S, siteyr.summ, 
         function(x) sum(!is.na(x))==0) |
  tapply(Data$TN.S, siteyr.summ, 
         function(x) sum(!is.na(x))==0)
siteyrsum.drop <- names(siteyrsum.drop)[siteyrsum.drop]

if (length(siteyrsum.drop)>0)
  Data <- Data[!is.element(grpyr.summ, grpyrsum.drop),]
```  

Step 5: Calculating Annual Geometric Means

- Annumal geometric means for each site

```{r}
#tapply(log(Data$TP.S), Data$site.yr, mean, na.rm=T)
```

- The result is put in a data frame named `TP.log`, along with site annual means of TN, CHLA, Turb, and other parameters

## Concluding Remarks

- Order of steps matters
- Several pages of code are difficult to debug
- Statistical concerns: geometric means are based on different sample sizes
